---
title: "mfrmr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
vignette_artifact <- function(name) {
  path <- system.file(
    "extdata", "vignette-artifacts", name,
    package = "mfrmr",
    mustWork = TRUE
  )
  read.csv(
    path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    na.strings = character(),
    colClasses = "character"
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

This vignette outlines a reproducible workflow for:

- loading packaged simulation data
- fitting an MFRM with flexible facets
- choosing a fast fit summary or an opt-in comprehensive summary
- producing the required Wright map on the fitted shared logit scale
- running diagnostics and residual PCA
- generating APA and visual summary outputs
- moving from fitted models into design simulation and fixed-calibration
  prediction

For a plot-first companion guide, see the separate
`mfrmr-visual-diagnostics` vignette.

For a faster preliminary run without changing the final analysis target:

- test code mechanics on a small deterministic subset or, for an MML workflow,
  a temporary `quad_points = 7` grid; restore the prespecified final MML grid
  (31 points by default) and check quadrature sensitivity before reporting
- choose `method = "JML"` only when its person-parameter treatment is
  methodologically appropriate, not merely as a faster substitute for MML
- use `diagnose_mfrm(..., residual_pca = "none", diagnostic_mode = "both",
  fit_df_method = "both")` when the diagnostics will feed a comprehensive
  summary
- reuse the same diagnostics object in downstream reports and plots

## MML and Diagnostic Modes

`mfrmr` treats `MML` and `JML` differently on purpose.

- `MML` integrates over the person distribution with Gauss-Hermite quadrature.
- `mml_engine = "direct"` (the default) optimizes the quadrature-based
  marginal log-likelihood directly. `mml_engine = "em"` and `"hybrid"` provide
  the documented EM and EM-warm-start routes for supported RSM/PCM fits.
- `JML` is useful for JMLE-oriented comparisons and analyses that avoid a
  parametric person distribution. `MML` is the package default and supports
  marginal and fixed-calibration follow-up when its response-model and
  population-distribution assumptions are defensible.

For `RSM` and `PCM`, diagnostics now expose two distinct evidence paths:

- `diagnostic_mode = "legacy"` keeps the residual/EAP-based stack.
- `diagnostic_mode = "marginal_fit"` adds the strict latent-integrated screen.
- `diagnostic_mode = "both"` is the safest default when you want to inspect
  both views side by side.

Strict marginal diagnostics are screening-oriented.
Use `summary(diag)$diagnostic_basis` to separate the legacy residual evidence
from the strict marginal evidence rather than pooling them into one decision.

## Load Data

```{r load-data}
library(mfrmr)

list_mfrmr_data(details = TRUE)[, c("Key", "PrimaryUse", "Design", "CountBasis")]

data("ej2021_study1", package = "mfrmr")
head(ej2021_study1)

study1_alt <- load_mfrmr_data("study1")
identical(names(ej2021_study1), names(study1_alt))
```

## Applied Runnable Example

Start with the packaged `example_operational` dataset. It is intentionally
compact but uses a connected two-rater assignment, unequal rater workloads,
and six planned omissions represented by absent long-format rows rather than
`NA` or sentinel scores. This makes the main tutorial closer to an applied
rating design without using empirical records. The same object is also
available via `data("mfrmr_example_operational", package = "mfrmr")`. A
separate score-free assignment roster lets the pre-fit review identify those
six omissions without guessing which cells should have existed.
`example_core` remains available as an idealized complete-crossing example for
fast help-page checks.

```{r toy-setup}
data("mfrmr_example_operational", package = "mfrmr")
data("mfrmr_example_operational_design", package = "mfrmr")
toy <- mfrmr_example_operational

data_review_toy <- describe_mfrm_data(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  rating_min = 1,
  rating_max = 4,
  expected_design = mfrmr_example_operational_design
)
data_summary_toy <- summary(data_review_toy)
data_summary_toy$structural_missingness
data_summary_toy$design_connectivity

fit_toy <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM"
)
diag_toy <- diagnose_mfrm(
  fit_toy,
  residual_pca = "none",
  diagnostic_mode = "both",
  fit_df_method = "both"
)

# Fast fit-only summary: this does not compute diagnostics.
fit_summary_toy <- summary(fit_toy, profile = "fit", detail = "brief")

# Comprehensive review, reusing diagnostics already computed above.
facets_summary_toy <- summary(
  fit_toy,
  profile = "facets",
  detail = "brief",
  diagnostics = diag_toy
)
res_toy <- facets_summary_toy$results

fit_summary_toy$overview
fit_summary_toy$readiness
fit_summary_toy$data_review
summary(diag_toy)$overview
facets_summary_toy

# Required first fitted-scale figure.
plot(res_toy, type = "wright", preset = "publication", show_ci = TRUE, top_n = Inf)

# Optional FACETS-style ruler. Replace these examples with the study rubric.
rubric_labels <- c(
  "1" = "Level 1",
  "2" = "Level 2",
  "3" = "Level 3",
  "4" = "Level 4"
)
plot(
  res_toy,
  type = "wright",
  renderer = "facets",
  category_labels = rubric_labels,
  show_ci = FALSE,
  preset = "publication"
)
# Setting show_ci = TRUE on this FACETS-style ruler is available as a
# deliberate hybrid: FACETS ruler grammar plus mfrmr uncertainty intervals.

# Optional follow-up: Infit on x, measure on y; persons are explicit opt-in.
plot(
  res_toy,
  type = "fit_pathway",
  fit_stat = "Infit",
  include_person = TRUE,
  top_n_person = 12,
  person_labels = "none",
  facet_labels = "flagged",
  preset = "publication"
)
```

Optimizer code zero is only one numerical signal. `fit_mfrm()` also checks the
terminal gradient; when the initial direct or hybrid solution stops with a
larger gradient and `reltol <= 1e-9`, it runs a bounded sequence of
warm-started polishing stages and retains the best non-worsening candidate
under the recorded selection rule. Inspect
`fit_toy$opt$optimizer_polish$Stages` when `Numerical` is not `pass`.

`maxit` is only a computational ceiling. It must not be increased until a
preferred coefficient, fit statistic, or expected conclusion appears. For a
final analysis, prespecify the estimator and controls and start from the
package default `maxit = 400` unless the analysis protocol states otherwise.
If a run reaches `ConvergenceStatus = "iteration_limit"`, keep it review-only
and repeat the same data, model, method, anchors, optimizer, tolerance, and
quadrature rule with the next ceiling in a prespecified sequence. Interpret
only a run with `Converged = TRUE`, `InferenceReady = TRUE`, and
`Numerical = pass`. If two separately ready runs differ materially, investigate
numerical stability instead of selecting the preferred result. Report the
requested ceiling, actual evaluations, convergence reason, and terminal
gradient.

`InferenceReady` describes this numerical gate only: Data, Design, Stability,
Diagnostics, and Reporting remain separate rows in `fit_summary_toy$readiness`.
For example, a disconnected design remains a reporting hold even if its
optimizer converges. Conversely,
`ready_for_diagnostics_and_reporting_follow_up` means that fitting completed
and diagnostic review is the next stage; it is neither an optimization failure
nor a claim that the analysis is already manuscript-ready.

`expected_design` is a declared roster, not an inferred complete crossing.
When no roster is available, the structural-missingness status is
`"not_declared"`; the package does not label unassigned cells as missing.

The `facets` profile is FACETS-style organization, not evidence that FACETS
was run and not a numerical-equivalence claim. Its brief print is selective
and does not print person identifiers; full tables remain in
`facets_summary_toy$results$tables`. For a report-oriented result set, use
`profile = "reporting"`. To inspect availability without allowing diagnostic
computation, use `compute = "never"`; requested dependent sections are then
recorded as `not_computed`:

```{r summary-profile-controls, eval = FALSE}
reporting_summary_toy <- summary(
  fit_toy,
  profile = "reporting",
  diagnostics = diag_toy
)

availability_only <- summary(
  fit_toy,
  profile = "facets",
  compute = "never"
)
availability_only$section_status
```

Availability and interpretability are intentionally separate. Review
`res_toy$readiness` and the `InterpretationStatus` columns in
`res_toy$plot_map` before treating an available plot as a final result.
Plots remain available during a numerical, data, design, or stability review,
but they warn and carry `REVIEW ONLY` in the returned subtitle and drawn title
until the readiness issue is resolved.

Bias/DIF, residual PCA, and anchor-drift/linking analyses are deliberately not
auto-run by any summary profile. They require explicit contrasts, diagnostic
settings, or multi-fit designs.

```{r toy-setup-artifacts, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_fit_overview.csv")
vignette_artifact("workflow_diagnostic_overview.csv")
vignette_artifact("workflow_plot_components.csv")
```

The same fit can then move through the recommended first reporting workflow:

```{r primary-route}
report_toy <- mfrm_report(res_toy, style = "qc")

summary(res_toy)$next_actions
summary(report_toy)$overview

# This is a controlled analysis archive, not a deidentified shareable export.
export_dir <- file.path(tempdir(), "mfrmr-workflow-export")
export_toy <- export_mfrm_results(
  res_toy,
  output_dir = export_dir,
  include = c("default", "report"),
  overwrite = TRUE,
  acknowledge_sensitive = TRUE
)
head(export_toy$written_files)
```

The acknowledgement suppresses the warning only; it does not redact person
identifiers, person-level results, local paths, or the complete RDS object.
Review every exported file under the study's data-handling policy before sharing.

```{r primary-route-artifacts, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_next_actions.csv")
vignette_artifact("workflow_report_overview.csv")
vignette_artifact("workflow_export_files.csv")
```

## Diagnostics and Reporting

```{r diagnostics-reporting}
t4_toy <- unexpected_response_table(
  fit_toy,
  diagnostics = diag_toy,
  abs_z_min = 1.5,
  prob_max = 0.4,
  top_n = 10
)
t12_toy <- fair_average_table(fit_toy, diagnostics = diag_toy)
t13_toy <- bias_interaction_report(
  estimate_bias(fit_toy, diag_toy,
                facet_a = "Rater", facet_b = "Criterion",
                max_iter = 2),
  top_n = 10
)

class(summary(t4_toy))
class(summary(t12_toy))
class(summary(t13_toy))

names(plot(t4_toy, draw = FALSE))
names(plot(t12_toy, draw = FALSE))
names(plot(t13_toy, draw = FALSE))

chk_toy <- reporting_checklist(fit_toy, diagnostics = diag_toy)
subset(
  chk_toy$checklist,
  Section == "Visual Displays",
  c("Item", "DraftReady", "NextAction")
)
```

```{r diagnostics-reporting-artifacts, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_summary_classes.csv")
vignette_artifact("workflow_plot_object_components.csv")
vignette_artifact("workflow_visual_checklist.csv")
```

## Fit and Diagnose with Full Data

For a larger sparse synthetic illustration, use the packaged Study 1 dataset:

```{r fit-full}
fit <- fit_mfrm(
  data = ej2021_study1,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)

diag <- diagnose_mfrm(
  fit,
  residual_pca = "none",
  diagnostic_mode = "both",
  fit_df_method = "both"
)

summary(fit, profile = "fit", detail = "brief")
summary(diag)

# Keep the final figure flow explicit: fit -> Wright map -> follow-up plots.
s <- summary(fit, profile = "facets", diagnostics = diag)
res <- s$results
s
plot(res, type = "wright", preset = "publication", show_ci = TRUE, top_n = Inf)

# Optional closest FACETS-style asterisk ruler (without mfrmr CI overlays).
plot(res, type = "wright", renderer = "facets",
     category_labels = rubric_labels, show_ci = FALSE,
     preset = "publication")

plot(
  res,
  type = "fit_pathway",
  fit_stat = "Infit",
  include_person = TRUE,
  top_n_person = 12,
  person_labels = "none",
  facet_labels = "flagged",
  preset = "publication"
)
```

This full-data figure caps the displayed person layer at 12 and suppresses
routine point labels to keep the first screen legible. The selected person IDs
and every retained facet row remain in
`plot(..., draw = FALSE)$data$table`; use `person_labels = "all"` or
`facet_labels = "all"` for a point-identification figure.

If you need residual-structure evidence for a final report, you can add residual
PCA after the initial diagnostic pass. Treat this as an exploratory screen, not
as a standalone unidimensionality test or as a DIMTEST/UNIDIM substitute. In
MFRM reporting, a cautious claim should combine global residual fit,
element-level fit, residual PCA, and local-dependence screens, for example:
"evidence consistent with essential unidimensionality under the specified facet
structure."

```{r fit-full-pca}
diag_pca <- diagnose_mfrm(
  fit,
  residual_pca = "both",
  pca_max_factors = 6
)

summary(diag_pca)
```

## Strict Diagnostics for RSM and PCM

For `RSM` and `PCM`, the package can now keep the legacy residual path and the
strict marginal path side by side:

```{r strict-rsm-pcm}
fit_rsm_strict <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7,
  maxit = 30
)
diag_rsm_strict <- diagnose_mfrm(
  fit_rsm_strict,
  diagnostic_mode = "both",
  residual_pca = "none"
)

fit_pcm_strict <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "PCM",
  step_facet = "Criterion",
  quad_points = 7,
  maxit = 30
)
diag_pcm_strict <- diagnose_mfrm(
  fit_pcm_strict,
  diagnostic_mode = "both",
  residual_pca = "none"
)

summary(diag_rsm_strict)$diagnostic_basis[, c("DiagnosticPath", "Status", "Basis")]
summary(diag_pcm_strict)$diagnostic_basis[, c("DiagnosticPath", "Status", "Basis")]
```

When you want a compact simulation-based screening check for the strict branch,
use `evaluate_mfrm_diagnostic_screening()` on a small design:

```{r strict-screening}
screen_rsm <- evaluate_mfrm_diagnostic_screening(
  design = list(person = 18, rater = 3, criterion = 3, assignment = 3),
  reps = 1,
  scenarios = c("well_specified", "local_dependence"),
  model = "RSM",
  maxit = 30,
  quad_points = 7,
  seed = 123
)
screen_pcm <- evaluate_mfrm_diagnostic_screening(
  design = list(person = 18, rater = 3, criterion = 3, assignment = 3),
  reps = 1,
  scenarios = c("well_specified", "step_structure_misspecification"),
  model = "PCM",
  maxit = 30,
  quad_points = 7,
  seed = 123
)

screen_rsm$performance_summary[, c("Scenario", "EvaluationUse", "LegacyAnyFlagRate", "StrictAnyFlagRate")]
screen_pcm$performance_summary[, c("Scenario", "EvaluationUse", "LegacySensitivityProxy", "StrictSensitivityProxy", "DeltaStrictMinusLegacyFlagRate")]
```

The same strict branch is now reflected in the reporting router:

```{r strict-reporting-route}
chk_rsm_strict <- reporting_checklist(fit_rsm_strict, diagnostics = diag_rsm_strict)
subset(
  chk_rsm_strict$checklist,
  Section == "Visual Displays" &
    Item %in% c("QC / facet dashboard", "Strict marginal visuals", "Precision / information curves"),
  c("Item", "Available", "DraftReady", "NextAction")
)
```

## Residual PCA and Reporting

```{r residual-pca}
pca <- analyze_residual_pca(diag_pca, mode = "both")
plot_residual_pca(pca, mode = "overall", plot_type = "scree")
```

```{r bias-apa}
data("mfrmr_example_bias", package = "mfrmr")
bias_df <- mfrmr_example_bias
fit_bias <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)
diag_bias <- diagnose_mfrm(fit_bias, residual_pca = "none")
bias <- estimate_bias(fit_bias, diag_bias, facet_a = "Rater", facet_b = "Criterion")
fixed <- build_fixed_reports(bias)
apa <- build_apa_outputs(fit_bias, diag_bias, bias_results = bias)

mfrm_threshold_profiles()
vis <- build_visual_summaries(fit_bias, diag_bias, threshold_profile = "standard")
vis$warning_map$residual_pca_overall
```

The same `example_bias` dataset also carries a `Group` variable so DIF-oriented
examples can show a non-null pattern instead of a fully clean result. It can be
loaded either with `load_mfrmr_data("example_bias")` or
`data("mfrmr_example_bias", package = "mfrmr")`.

## Human-Readable Reporting API

```{r reporting-api}
spec <- specifications_report(fit, title = "Study run")
data_qc <- data_quality_report(
  fit,
  data = ej2021_study1,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score"
)
iter <- estimation_iteration_report(fit, max_iter = 8)
subset_rep <- subset_connectivity_report(fit, diagnostics = diag)
facet_stats <- facet_statistics_report(fit, diagnostics = diag)
cat_structure <- category_structure_report(fit, diagnostics = diag)
cat_curves <- category_curves_report(fit, theta_points = 101)
bias_rep <- bias_interaction_report(bias, top_n = 20)
plot_bias_interaction(bias_rep, plot = "scatter")
```

## Design Simulation and Prediction

The package also supports a separate simulation/prediction layer. The key
distinction is:

- `evaluate_mfrm_recovery()` checks whether known generating parameters are
  recovered under a stated simulation design. It is the first simulation check
  to run when you are validating a model specification or a planned design.
- `evaluate_mfrm_design()` and `predict_mfrm_population()` are
  design-level helpers that summarize expected operating characteristics under
  an explicit simulation specification.
- `mfrm_generalizability()` and `mfrm_d_study()` summarize observed
  univariate G-study components and analytic D-study projections. Read
  `IdentificationStatus`, `GStatus`, and `PhiStatus` before reporting
  projected coefficients; boundary or singular mixed-model fits are
  design-identification warnings rather than high-stakes-ready reliability
  evidence.
- `predict_mfrm_units()` and `sample_mfrm_plausible_values()` score future or
  partially observed persons under a fixed `MML` calibration.

```{r gd-study-reading-order, eval=FALSE}
if (requireNamespace("lme4", quietly = TRUE)) {
  gt <- mfrm_generalizability(fit)
  gt$coefficients[, c("G", "Phi", "GStatus", "PhiStatus",
                      "IdentificationStatus")]

  ds <- mfrm_d_study(
    gt,
    data.frame(Rater = c(2, 3, 4), Criterion = 4),
    residual_scaling = "sensitivity"
  )
  ds[, c("n_Rater", "n_Criterion", "ResidualScaling",
         "G", "Phi", "GStatus", "PhiStatus", "IdentificationStatus")]
}
```

```{r design-prediction}
sim_spec <- build_mfrm_sim_spec(
  n_person = 30,
  n_rater = 4,
  n_criterion = 4,
  raters_per_person = 2,
  assignment = "rotating"
)

recovery <- suppressWarnings(
  evaluate_mfrm_recovery(
    sim_spec = sim_spec,
    reps = 2,
    maxit = 30,
    include_diagnostics = TRUE,
    diagnostic_fit_df_method = "both",
    seed = 2
  )
)

summary(recovery)$recovery_summary[, c("ParameterType", "Facet", "RMSE", "Bias")]
plot(recovery, type = "summary", metric = "rmse", draw = FALSE)$data$plot_table

recovery_review <- assess_mfrm_recovery(
  recovery,
  min_reps = 2,
  min_se_available = NULL,
  max_mcse_rmse_ratio = NULL,
  max_rmse = c(facet = 1, step = 1, default = 1.5),
  max_abs_bias = c(default = 0.75)
)

summary(recovery_review)$checklist[, c("Section", "Item", "Status")]
summary(recovery_review)$reading_order
summary(recovery_review)$condition_reporting_notes
summary(recovery_review)$condition_review
summary(recovery_review)$diagnostic_reporting_notes
summary(recovery_review)$diagnostic_review

status_plot <- plot(recovery_review, type = "status", draw = FALSE)
status_plot$data$section_status
status_plot$data$reading_order

metric_plot <- plot(recovery_review, type = "metrics", metric = "rmse", draw = FALSE)
metric_plot$data$plot_table
metric_plot$data$guidance

recovery_bundle <- build_summary_table_bundle(
  recovery_review,
  appendix_preset = "recommended"
)
recovery_bundle$table_index[, c("Table", "Rows", "Role")]

pred_pop <- predict_mfrm_population(
  sim_spec = sim_spec,
  reps = 2,
  maxit = 30,
  seed = 1
)

summary(pred_pop)$forecast[, c("Facet", "MeanSeparation", "McseSeparation")]

keep_people <- unique(toy$Person)[1:18]
toy_mml <- suppressWarnings(
  fit_mfrm(
    toy[toy$Person %in% keep_people, , drop = FALSE],
    person = "Person",
    facets = c("Rater", "Criterion"),
    score = "Score",
    method = "MML",
    quad_points = 5,
    maxit = 30
  )
)

new_units <- data.frame(
  Person = c("NEW01", "NEW01"),
  Rater = unique(toy$Rater)[1],
  Criterion = unique(toy$Criterion)[1:2],
  Score = c(2, 3)
)

pred_units <- predict_mfrm_units(toy_mml, new_units, n_draws = 0)
pv_units <- sample_mfrm_plausible_values(toy_mml, new_units, n_draws = 2, seed = 1)

summary(pred_units)$estimates[, c("Person", "Estimate", "Lower", "Upper")]
summary(pv_units)$draw_summary[, c("Person", "Draws", "MeanValue")]
```

For a report or appendix handoff, pass the recovery objects through the same
summary-table export route used by the rest of the package:

```{r recovery-export, eval = FALSE}
export_summary_appendix(
  list(recovery = recovery, recovery_review = recovery_review),
  output_dir = tempdir(),
  prefix = "mfrmr_recovery_appendix",
  preset = "recommended",
  include_html = FALSE,
  overwrite = TRUE
)
```

For an initial exploratory run, `reps = 2` or another very small value is useful
only to check the data-generating setup and refit path. For a study report,
increase `reps`, keep the ADEMP-style metadata in the exported tables, and set
substantive RMSE/Bias thresholds so that `assess_mfrm_recovery()` can mark those
rows as `ok`, `review`, or `concern` rather than `not_assessed`.
