---
title: "mfrmr Visual Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Visual Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette is a compact map of the main base-R diagnostics in `mfrmr`.
It is organized around four practical questions:

- How well do persons, facet levels, and categories target each other?
- Which observations or levels look locally unstable?
- Is the design linked well enough across subsets or forms?
- Where do residual structure and interaction screens point next?

All examples use packaged data and `preset = "publication"` so the same code
is suitable for manuscript-oriented graphics.

If you are selecting figures for a report, use `reporting_checklist()` before
or alongside this vignette. Its `"Visual Displays"` rows now mirror the
public plotting family shown here.

## Minimal setup

```{r setup}
library(mfrmr)

toy <- load_mfrmr_data("example_core")

fit <- fit_mfrm(
  toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "JML",
  model = "RSM",
  maxit = 20
)

diag <- diagnose_mfrm(fit, residual_pca = "none")
checklist <- reporting_checklist(fit, diagnostics = diag)
subset(
  checklist$checklist,
  Section == "Visual Displays",
  c("Item", "Available", "NextAction")
)
```

## 1. Targeting and scale structure

Use the Wright map first when you want one shared logit view of persons,
facet levels, and step thresholds.

```{r wright}
plot(fit, type = "wright", preset = "publication", show_ci = TRUE)
```

Interpretation:

- Compare person density on the left to facet and step locations on the right.
- Large gaps suggest weaker targeting in that logit region.
- Wide overlap in confidence whiskers means neighboring levels are not cleanly separated.

Next, use the pathway map when you want to see how expected scores progress
across theta.

```{r pathway}
plot(fit, type = "pathway", preset = "publication")
```

Interpretation:

- Steeper rises indicate stronger score progression.
- Dominant-category strips show where each category is most likely to govern the score.
- Flat or compressed regions suggest weaker category separation.

## 2. Local response and level issues

Unexpected-response screening is useful for case-level review.

```{r unexpected}
plot_unexpected(
  fit,
  diagnostics = diag,
  abs_z_min = 1.5,
  prob_max = 0.4,
  plot_type = "scatter",
  preset = "publication"
)
```

Interpretation:

- Upper corners combine large residual mismatch with low model probability.
- Repeated appearances of the same persons or levels are more informative than a single extreme point.

Displacement focuses on level movement rather than individual responses.

```{r displacement}
plot_displacement(
  fit,
  diagnostics = diag,
  anchored_only = FALSE,
  plot_type = "lollipop",
  preset = "publication"
)
```

Interpretation:

- Large absolute displacement indicates stronger tension between observed data and current calibration.
- For anchored runs, this is especially useful as an anchor-robustness screen.

### Strict marginal follow-up

When you need the package's latent-integrated follow-up path, switch to `MML`
and request `diagnostic_mode = "both"` so the legacy and strict branches stay
visible side by side.

```{r strict-marginal}
fit_strict <- fit_mfrm(
  toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7,
  maxit = 40
)

diag_strict <- diagnose_mfrm(
  fit_strict,
  residual_pca = "none",
  diagnostic_mode = "both"
)

strict_checklist <- reporting_checklist(fit_strict, diagnostics = diag_strict)
subset(
  strict_checklist$checklist,
  Section == "Visual Displays" &
    Item %in% c("QC / facet dashboard", "Strict marginal visuals"),
  c("Item", "Available", "NextAction")
)

plot_marginal_fit(
  diag_strict,
  top_n = 12,
  preset = "publication"
)
```

Interpretation:

- Treat strict marginal plots as exploratory corroboration screens, not as standalone inferential tests.
- Use the checklist rows to confirm that the current run actually supports the strict branch before routing figures into a report.
- When pairwise follow-up is needed, continue with `plot_marginal_pairwise(diag_strict, preset = "publication")`.

## 3. Linking and coverage

When the design may be incomplete or spread across subsets, inspect the
coverage matrix before interpreting cross-subset contrasts.

```{r linking}
sc <- subset_connectivity_report(fit, diagnostics = diag)
plot(sc, type = "design_matrix", preset = "publication")
```

Interpretation:

- Sparse rows or columns indicate weak subset coverage.
- Facets with low overlap are weaker anchors for cross-subset comparisons.

If you are working across administrations, follow up with anchor-drift plots:

```{r eval = FALSE}
drift <- detect_anchor_drift(current_fit, baseline = baseline_anchors)
plot_anchor_drift(drift, type = "heatmap", preset = "publication")
```

## 4. Residual structure and interaction screens

Residual PCA is a follow-up layer after the main fit screen.

```{r residual-pca}
diag_pca <- diagnose_mfrm(fit, residual_pca = "both", pca_max_factors = 4)
pca <- analyze_residual_pca(diag_pca, mode = "both")
plot_residual_pca(pca, mode = "overall", plot_type = "scree", preset = "publication")
```

Interpretation:

- Early components with noticeably larger eigenvalues deserve follow-up.
- Scree review should usually be paired with loading review for the component of interest.

For interaction screening, use the packaged bias example.

```{r bias}
bias_df <- load_mfrmr_data("example_bias")

fit_bias <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)

diag_bias <- diagnose_mfrm(fit_bias, residual_pca = "none")
bias <- estimate_bias(fit_bias, diag_bias, facet_a = "Rater", facet_b = "Criterion")

plot_bias_interaction(
  bias,
  plot = "facet_profile",
  preset = "publication"
)
```

Interpretation:

- Facet profiles are useful for seeing whether a small number of levels drives most flagged interaction cells.
- Treat these plots as screening evidence; confirm with the corresponding tables and narrative reports.

## Recommended sequence

For a compact visual workflow:

1. `reporting_checklist()` when you want the package to route which figures are already supported.
2. `plot_qc_dashboard()` for one-page triage.
3. `plot_unexpected()`, `plot_displacement()`, `plot_marginal_fit()`, and `plot_interrater_agreement()` for local follow-up.
4. `plot(fit, type = "wright")` and `plot(fit, type = "pathway")` for targeting and scale interpretation.
5. `plot_residual_pca()`, `plot_bias_interaction()`, and `plot_information()` for deeper structural review.

## Related help

- `help("mfrmr_visual_diagnostics", package = "mfrmr")`
- `help("mfrmr_workflow_methods", package = "mfrmr")`
